\documentclass[a4paper,12pt]{article}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{txfonts}
%\usepackage{upgreek}
\usepackage{graphicx}
%\usepackage{siunitx}
\usepackage{enumerate}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}

%\newcommand{\question}[2]{\textbf{\textit{#1}}\quad{\footnotesize\textit{(#2 points)}}\\[3mm]}
\newcommand{\question}[1]{\textbf{\textit{#1}}}
\newcommand{\points}[1]{\quad{\footnotesize\textit{(#1 points)}}}
\newcommand{\point}{\quad{\footnotesize\textit{(1 point)}}}
\newcommand{\HRule}{\rule{\linewidth}{0.3mm}}
\newcommand{\dd}{\mathrm{d}}
\renewcommand{\pi}{\uppi}
\newcommand{\ii}{\mathrm{i}}
\renewcommand{\thefootnote}{\normalsize\fnsymbol{footnote}}
\DeclareMathOperator{\e}{e}

\renewcommand{\theequation}{\Roman{equation}}

\begin{document}
\pagestyle{empty}

\begin{center}
\LARGE \textbf{Astronomy from 4 perspectives: the Dark Universe}
\HRule
\end{center}
\begin{flushright}
prepared by: Florence participants
\end{flushright}
\begin{center}
{\Large \textbf{questions: Supernova-cosmology and dark energy}}
\end{center}
\vspace{5mm}

\begin{enumerate}[\itshape \bfseries 1.]

\item \question{FLRW-models and the equation of state}\\
\begin{enumerate}[(a)]
\item{What are the ingredients (fields?) the enter into a FLRW
    metric?}
\item{How does each field behave as the universe expands?}
\end{enumerate}

\item \question{light propagation in relativity}\\
\begin{enumerate}[(a)]
\item{What is the equation that describes light propagation in a
    curved space-time?}
\item{How does the presence of a body exerting gravity affect light
    paths?}
\item{How curvature affect light propagation?}
\item{How expansion affects light propagation?}
\end{enumerate}

\item \question{distance measures}\\
\begin{enumerate}[(a)]
\item{How do we measure distance in a curved space-time?}
\item{What is the angular distance?}
\item{How we disentangle red-shift from curvature?}
\end{enumerate}

\item \question{supernova-cosmology}\\
\begin{enumerate}[(a)]
\item{Why SN IA are standard candles?}
\item{Why is it important to observe them at high redshift?}
\item{Why are they not enough to fully constrain the cosmology?}
\end{enumerate}


\end{enumerate}
\end{document}
