\documentclass[a4paper,12pt]{article}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{txfonts}
\usepackage{upgreek}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage{enumerate}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}

%\newcommand{\question}[2]{\textbf{\textit{#1}}\quad{\footnotesize\textit{(#2 points)}}\\[3mm]}
\newcommand{\question}[1]{\textbf{\textit{#1}}}
\newcommand{\points}[1]{\quad{\footnotesize\textit{(#1 points)}}}
\newcommand{\point}{\quad{\footnotesize\textit{(1 point)}}}
\newcommand{\HRule}{\rule{\linewidth}{0.3mm}}
\newcommand{\dd}{\mathrm{d}}
\renewcommand{\pi}{\uppi}
\newcommand{\ii}{\mathrm{i}}
\renewcommand{\thefootnote}{\normalsize\fnsymbol{footnote}}
\DeclareMathOperator{\e}{e}

\renewcommand{\theequation}{\Roman{equation}}

\begin{document}
\pagestyle{empty}

\begin{center}
\LARGE \textbf{Astronomy from 4 perspectives: the Dark Universe}
\HRule
\end{center}
\begin{flushright}
prepared by: Heidelberg participants
\end{flushright}
\begin{center}
{\Large \textbf{questions: Dark matter and the virial theorem}}
\end{center}
\vspace{5mm}

\begin{enumerate}[\itshape \bfseries 1.]

\item \question{theory behind the virial theorem}
\begin{enumerate}[(a)]
\item{What's the idea of the virial theorem?}
\item{What's the difference to energy conservation?}
\item{For what kind of system can you use the virial theorem?}
\end{enumerate}

\item \question{mechanical similarity}
\begin{enumerate}[(a)]
\item{Why is mechanical similarity restricted to potentials of the shape $\Phi\propto r^\alpha$?}
\item{Does $\alpha$ have to be integer?}
\item{What's the generalisation of the Kepler-law for a potential of the form $\Phi\propto r^\alpha$?}
\end{enumerate}

\item \question{harmonic oscillator}
\begin{enumerate}[(a)]
\item{What can you say about different energy types in the harmonic oscillator on average?}
\end{enumerate}

\item \question{Kepler-problem and planetary motion}
\begin{enumerate}[(a)]
\item{Can you show that Kepler's third law implies that potential and kinetic energy are proportional to each other?}
\item{Is this true in general?}
\end{enumerate}

\item \question{virial theorem in galaxies}\\
\begin{enumerate}[(a)]
\item{Could one explain the rotation curves by assuming a different gravitational law?}
\end{enumerate}

\end{enumerate}
\end{document}
